\large{\bf Looking forwards and backwards through locally regulated populations }

### 2023-01-28 forwards-and-backwards.tex
alison rewrite through p.65

\documentclass[12pt]{article}
\usepackage{fullpage}
\usepackage{amssymb}
\usepackage{graphics} 
\usepackage{amsmath}
\usepackage[color=yellow]{todonotes}
\usepackage{url}
\usepackage{xcolor}
\usepackage{longtable}
\usepackage[hidelinks]{hyperref}
\usepackage{bbm}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{hypothesis}[theorem]{Hypothesis}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{example}[theorem]{Example}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{notation}[theorem]{Notation}

@@

% OTHER IDEAS:
% A class of nonlinear superprocesses
% Uncovering genealogical structures of locally regulated populations with lookdown constructions on nonlinear superprocesses

@@

We do something...

make sure to put in here that we get convergence of the FKPP

Note: use "reproductive value" instead of "long-term fitness"
and say clearly that "lineage" means "ancestral lineage".

@@

\noindent {\bf Key words:}  population model, nonlinear superprocess, 
lookdown construction, porous medium equation, 
reaction-diffusion equation, travelling waves, genealogies,

@@

%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

\comment{I think the introduction covers a good set of things in a good order, it's just a bit long.}

\comment{
    Make sure that we discuss the difference to Flandoli's model
    (one of the few others that show convergence of FKPP),
    and say that a novelty of our results is that we show convergence
    for FKPP and broader models.
}

@@

As one takes a journey, long or short, the landscape changes:
forests thicken or thin or change their composition,
springtime grasslands host intergrading mosaics of different types of flowers.
The aim of this paper is to understand the implications of a broad class of mechanistic spatial models
that might describe how such spatially heterogeneous populations live, die, and reproduce:
How does population density change across space and time?
How might we learn about the underlying dynamics from genealogical or genetic data?
And,  how does genetic ancestry spread across geography,
when looking back through time in these populations?
We work with individual-based models of a single species in continuous space,
where birth, death and establishment may all depend on spatial location
as well as on local population density, allowing for stable populations through density-dependent feedback.

Reproduction of individuals naturally leads to a branching process.
There is now a huge literature devoted to spatial branching processes,
including branching random walk, 
branching Brownian motion, and the Dawson-Watanabe superprocesses.  
In addition to possessing a rich and beautiful mathematical structure, 
the superprocesses are remarkable for demonstrating a certain `universality', 
arising, as they do, as scaling limits of a vast array of different 
spatial population models \cite[e.g.,][]{chetwynd-diggle/etheridge:2018,
cox/perkins:2005,
cox/durrett/perkins:1999,
cox/durrett/perkins:2000,
holmes:2008,
vanderhofstad/sakai:2010,
vanderhofstad/slade:2003,
vanderhofstad/holmes/perkins:2017}.

These models make mathematical sense in very general spaces, but
in biological applications, typically, 
individuals are assumed to be scattered across one or two-dimensional
Euclidean space. In that setting, it is well known that, 
in the long term, the population will either 
die out, or it will develop clumps of arbitrary density and extent. 
To counter this, it is common to
impose some local regulation by either increasing death rates, or 
reducing birth rates, in proportion to the local population density,
so that the population has a tendency to grow in sparsely populated regions
and shrink where it is overcrowded. This leads to spatial analogues of  
logistic branching processes.

Although it can be more mathematically convenient to work with branching
Brownian motions, in many settings it is biologically more natural to 
develop models based on branching random walk.  
Individuals produce a random number of offspring,
that are thrown off according to some (usually symmetric) 
distribution centred on the location of the parent.   
This is particularly appropriate for modelling plant populations, in which
this dispersal of offspring around the parent is the only source of
spatial motion.

Most models do not distinguish between juveniles and adults, so,
for example, the number of adults produced by a single parent is determined
by the degree of crowding at the location of the parent. The novelty
of the models that we introduce here is that, although we shall only
follow the adult population, in formulating the dynamics of the
models we shall distinguish
between production of juveniles, which will depend upon the location of 
the adult, and their successful establishment, which will depend on the
location in which a juvenile lands. The result is that not only the absolute 
number, but also the spatial distribution
around their parent, 
of those offspring that survive to adulthood
will depend upon the local population 
density.

We shall consider different classes of scaling limits for our model; the first class consisting of
(generalised) super-processes limits, and the second class where the limits are deterministic. In the
superprocess setting, for our approach to work we measure the local population
density at a point by integrating against a smooth test function, for example 
a Gaussian density centred on that point.  
When the limit is deterministic,
we can simultaneously reduce the parameter in that Gaussian density so that,
in the limit, the local population density is simply the population density 
at that point. Nonetheless, we retain a signal of our two stage approach
to modelling reproduction, and  
in this way we recover a nonlinear partial differential equation. 
This approach requires some technical conditions that we will verify in two important examples,
the porous medium equation with a logistic growth term of the form 
$$\partial_t \varphi = \Delta (\varphi^2)+\varphi(1-\varphi),$$
and a wide class of semi-linear partial differential equations of the form 
$$\partial_t \varphi = \Delta\varphi+ \varphi \left(G(\varphi)-H(\varphi)\right),$$
which include the Fisher-KPP equation and the Allen-Cahn equation.

The porous medium equation (PME)
is a degenerate nonlinear diffusion equation that
has been the subject of intensive study. We refer to~\cite{vazquez:2007} for 
a comprehensive reference. The equation has also been studied with various
forms of noise, see~\cite{barbu/daprato/roeckner:2016}, although not the 
form of noise that will arise from the population dynamics below. 
Deriving nonlinear diffusion equations as limits of weakly interacting 
diffusions goes back at least to~\cite{mckean:1967}, and this 
programme has been
carried out for the PME and its generalisations. 
The degeneracy of the PME
results in some extra analytic
challenges,~\cite{jourdain:2000}.
% Recent work in this 
% direction has been spurred on by applications
% in mathematical finance, %where the porous medium arises 
% in particular, in the study of weak limits of systems
% of diffusions interacting through their 
% rank,~\cite{dembo/shkolnikov/varadhan/zeitouni:2016, jourdain/reygner:2013}
% and references therein.
Our approach is different.
We start from individual
based population models, thus providing a microscopic mechanism that leads to 
nonlinear diffusion.
Other microscopic mechanisms have been proposed in the special case of the
PME, see 
e.g.~\cite{feng/iscoe/seppalainen:1997, oelschlaeger:1990}; ours
is specifically motivated by biological considerations.

Another novelty of our models stems from the use of lookdown construction to retain the dynamics of lines of descent in our population while passing to a scaling limit. The lookdown construction is first introduced in \cite{donnelly/kurtz:1996} to provide a mechanism to trace lineages while passing through a scaling limit for the Moran model. In more recent works, see~\cite{kurtz/rodrigues:2011, etheridge/kurtz:2018}, lookdown constructions are further applied into models of branching populations with spatial structures. In \S \ref{Sec: Lookdown Constructions}, we follow closely the techniques presented in these papers to impose lookdown constructions for our models which allow us to retain information about ancestry of individuals in the population as we pass to a large population limit. This in turn allows us to write down an explicit expression for the generator of the position of ancestral lineages with respect to the travelling wave fronts for populations modelled by a huge class of semi-linear partial differential equations, see Theorem \ref{teo:ancestral lineages}.
This attempt is novel and we have provided the first ever rigorous proof to similar results on ancestral lineages. This result allows us to compare ancestral lineages of individuals sampled from the front of a population evolving according to the PME with logistic growth to those under the Fisher-KPP and Allen-Cahn equations. For example, as a result of the nonlinear diffusion, 
for suitable initial conditions, the solution to the one-dimensional 
porous medium equation with logistic growth converges to a travelling
wave with a sharp cut-off; i.e., in contrast to the classical 
Fisher KPP equation, the solution at time $t$
vanishes beyond $x=x_0+ct$ for some constant wavespeed
$c>0$, \cite{kamin/rosenau:2004}.

\comment{Need a transition or otherwise to merge this with the above.}
The history of a natural population is often only accessible indirectly 
through patterns of genetic diversity that have been laid down. From 
genetic data, one can try to infer the genealogical trees that relate 
individuals in a sample from the population. It is therefore of interest
to establish the distribution of genealogical trees relating individuals
sampled from a population evolving according to any proposed mathematical
model.

In general this is an extremely difficult question. However, in special
circumstances, some progress can be made. One of the settings that
has received a great deal of attention in recent years is that in which 
a population is expanding into new territory as a travelling wave. 
Typically one considers a stationary wavefront moving across $\IR^1$, and
most work has focussed on the classical Fisher-KPP equation with a 
stochastic term, i.e.
$$dw=\big(\Delta w +sw(1-w)\Big)dt +\sqrt{\frac{\alpha(w)}{N}}W(dt,dx),$$
where $W$ is space-time white noise, and $N$ is a measure of the 
local population density. The coefficient
$\alpha(w)$ is generally taken to be either
$w$, corresponding to a superprocess limit, or $w(1-w)$ giving a 
spatial analogue of a Wright-Fisher diffusion. The first is 
appropriate for modelling the range of an expanding population,
the second for the spread of a selectively advantageous 
mutation through an established (selectively neutral) population. 
Because of the logistic control of the growth rate,
individuals in the wavefront can have much greater reproductive success
than those in the `bulk' of the expanding population. 
Tracing backwards in time, individuals sampled
from behind the wave are `caught' by the wave, upon which they become
trapped in the wavefront. If the population density is very high, then over
suitable timescales, the genealogical trees will be dominated by 
periods of extremely rapid coalescence corresponding to a significant 
proportion of the individuals in the front being descended from a particularly
reproductively successful ancestor. For a surprisingly wide range of models, 
it is believed that, for very large $N$ and on suitable timescales, 
genealogies converge to a 
Bolthausen-Sznitman coalescent, e.g.~\cite{brunet/derrida/mueller/munier:2007}. 
On the other hand, if one replaces the logistic growth term of the classical
Fisher-KPP equation with a nonlinearity that reflects cooperative
behaviour in the population, such as
$$wF(w)=w(1-w)(Cw-1),$$
then, for sufficiently large $C$ (strong cooperation),
the nature of the deterministic
wave changes from `pulled' to `pushed', \cite{birzu/hallatschek/korolev:2017}.
In that setting, the genealogies will also be quite different
from the Fisher-KPP case. For example, \cite{etheridge/penington:2020}
shows that for a discrete space model corresponding to this 
nonlinearity with $C>4$, after suitable scaling, the genealogy of a
sample converges not to a Bolthausen-Sznitman coalescent, but to
a Kingman coalescent. 
\todo[inline]{Is that condition $C>4$ correct? the paper with Sarah is done for the equation $\partial_t u = \Delta u + u(1-u)(2u-1+\alpha)$ and valid for $\alpha \in (0,1)$. Factorising we would get that equation would be $\partial_t u = \Delta u + (1-\alpha) u(1-u)(\frac{2}{1-\alpha}u-1)$. Supposing the $(1-\alpha)$ in front does not affect the result (which I suppose wouldn't) we get that $C= \frac{2}{1-\alpha}$ and so $\alpha \in (0,1)$ if and only if $C>2$ ...}
The reason, roughly, is that ancestral lineages
settle to a stationary distribution relative to the position of the 
wavefront which puts very little weight close to the `tip' of the wave, so
that when ancestral lineages meet,
it is typically in a location where population density is
high, where no single ancestor produces a disproportionately large number of 
descendants in a short space of time.

As a first step towards understanding what we should expect in models with
nonlinear diffusion, we consider the position of an ancestral lineage
relative to the wavefront in the deterministic models, and compare that 
to what we see for a classical Fisher-KPP equation. 
In~\S\ref{sec: wright-malecot}, we then indicate how to extend these ideas to 
extract information about relatedness between two individuals in the 
superprocess limit, at least when population intensity is very high, and
we present some evidence to support the conjecture that genealogies 
for populations expanding according to 
our superprocess models with nonlinear diffusion
can be very different from those under 
the classical Fisher-KPP equation.

% % % % % % % % % % % % % %
\paragraph{Structure of the paper}
In the paper we study scaling limits of spatial populations,
obtaining convergence of both the population process
(i.e., the population density as a function of time,
although strictly speaking it is a measure that may not have a density)
as well as of lineages traced back through such a population.
We retain linages through the rescaling process
by means of a lookdown construction.
First, in section~\ref{sec: Model and main results},
we describe the model and the main theorems
(Theorems~\ref{thm:nonlocal_convergence}, \ref{thm:local_convergence}, and~\ref{thm:lineages}).
Next, in Section~\ref{sec:applications}, we discuss a few striking consequences of these results,
namely, X Y and Z \comment{TODO}.
In section \ref{sec:heuristics}, we provide heuristic explanations
of why the theorems ought to be true,
and some key ideas behind them.
In Section \ref{sec:lookdown} we define and discuss the lookdown process that retains the levels.
This is followed by the technical proofs in section \ref{sec:proofs}.

@@

The model we work with is a fairly general model
of individuals living in continuous space.
We start with a population of
$\bigO(N)$ individuals per unit area distributed in $\IR^d$.
For applications, $d=1$ or $d=2$, but our results apply more generally.
The population will be distributed over a bounded region,
so the total number of individuals will also be $\bigO(N)$.
The population changes in continuous time, and
we describe the state of the population at time $t$ with a counting measure $X(t)$,
which assigns one unit of mass to the location of each individual.

Population dynamics are controlled by three quantities,
birth ($\gamma$), establishment ($r$), and death ($\mu$),
each of which can depend on spatial location and local population density
in a way we specify later.
Each individual gives birth at rate $\gamma$ to a single offspring,
and the offspring
then disperses according to a kernel $q(\cdot)$ away from the location of the parent.
We assume that $q$ is the density of a multivariate Gaussian,
allowing a nonzero mean and anisotropic variance.
Both the mean and covariance can change across space,
but do not depend on population density.
The offspring does not necessarily survive to be counted in the population:
it ``establishes'' with probability $r$,
or else it dies immediately.
Independently, each individual dies with rate $\mu$.

We aim to obtain universal behavior by taking the density, $N$, to infinity,
and also scaling time by a factor of $\theta$,
so that defining $\eta(t) = X(\theta t) / N$,
the process $\{\eta(t)\}_{t \ge 0}$
will converge to a suitable measure-valued process
as $N$ and $\theta$ tend to infinity,
with the nature of the limit depending on how they tend to infinity together.
Demographic rates in our model
will be determined by local population density relative to $N$,
so that population dynamics depend more naturally on $\eta$ (rather than $X$).
However, to avoid confusion arising from the time scaling,
we describe the model in terms of $X$ next.
See Figure~\ref{fig:model_setup} for a conceptual depiction.

\begin{figure}
    \begin{center}
        \includegraphics{figures/conceptual_figure}
    \end{center}
    \caption{
        \comment{Perhaps we should have two panels, one with the PME example (currently panel A) and the other with FKPP;
        right now panel B is not very helpful I think, so maybe should be omitted or added to so it better describes the whole process
        (including dispersal and establishment).}
        Conceptual figure of the model.
        \textbf{(A)} Example plot of $\gamma$, $\mu$, and $F$
        (calculated here with $r=1$) against scaled local population density.
        In this example, both birth and death rates increase with population density,
        but death increases faster,
        and equilibrium occurs where they cross
        (and hence where $F=0$).
        \textbf{(B)} Cartoon of interaction ($\epsilon$) and dispersal ($\sigma$) distances.
        \label{fig:model_setup}
    }
\end{figure}

>>>>

\documentclass[12pt]{article}
\usepackage{fullpage}
\usepackage{amssymb}
\usepackage{ntheorem}
\usepackage{graphics} 
\usepackage{amsmath}
\usepackage[color=yellow]{todonotes}
\usepackage{url}
\usepackage{xcolor}
\usepackage{longtable}
\usepackage[hidelinks]{hyperref}
\usepackage{bbm}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{hypothesis}[theorem]{Hypothesis}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{example}[theorem]{Example}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{assumptions}[theorem]{Assumptions}
\newtheorem{terminology}[theorem]{Terminology}
\newtheorem*{terminology-non}[theorem]{Terminology}

@@

@@

We introduce a broad class of mechanistic spatial models to describe how spatially 
heterogeneous populations live, die, and reproduce. Birth and death rates can depend 
both on spatial position and local population density, defined at a 
spatial position $x$ to be the convolution 
of the point measure encoding the distribution of individuals in the population
with a suitable non-negative integrable kernel
centred on $x$. We pass to three different scaling limits:
an interacting superprocess, a nonlocal partial differential equation (PDE), 
and a classical PDE. 
The classical PDE is obtained both by a two-step convergence argument, in which we 
first scale time and population size and
pass to the nonlocal PDE, 
and then scale the kernel that 
determines local population density; and 
in the important special case in which the limit is a 
reaction diffusion equation, directly 
by simultaneously scaling the kernel and the
timescale and population size in our individual based model.

A novelty of our model is that in describing the dynamics we explicitly
model a juvenile phase: the number of juveniles produced by an individual depends
on local population density at the location of the parent. 
Juvenile offspring are 
thrown off in a (possibly heterogeneous, anisotropic) Gaussian distribution 
around the location of the parent; they then reach (instant) maturity with a probability
that can depend on the local population density (possibly measured through convolution
with a different kernel from that used in determining fecundity) 
at the location at which they land. 
Although we only record mature individuals, a trace of this two-step description 
remains in our population models, resulting in novel limits in which the spatial
motion is governed by a nonlinear diffusion.

Using a lookdown representation, we are able to retain information about 
genealogies relating individuals in our population and, in the case of 
deterministic limiting models, we use this to deduce 
the backwards in time motion of the ancestral lineage of an individual
sampled from the population. We observe that
knowing the history of the population density is not enough to determine the motion of
ancestral lineages in our model. We also investigate (and
contrast) the behaviour of lineages for 
three different deterministic models of a population expanding its range as a 
travelling wave: the Fisher-KPP equation, the Allen-Cahn equation, and a porous 
medium equation with logistic growth.

@@

	\comment{Need to thank KAVLI, Paris Cit\'e, various funders}

\noindent {\bf Key words:}  population model, interacting superprocess, 
lookdown construction, porous medium equation, 
reaction-diffusion equation, travelling waves, genealogies,
Fisher-KPP equation

@@

%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}

@@

As one takes a journey, long or short, the landscape changes:
forests thicken or thin or change their composition;
springtime grasslands host intergrading mosaics of different types of flowers.
The aim of this paper is to introduce and study
a broad class of mechanistic spatial models
that might describe how spatially heterogeneous populations live, die, and reproduce.
Questions that we (start to) address include:
How does population density change across space and time?
How might we learn about the underlying dynamics from genealogical or genetic data?
And, how does genetic ancestry spread across geography
when looking back through time in these populations?

Reproduction of individuals naturally leads to spatial branching process models,
including branching random walk, 
branching Brownian motion, and the Dawson-Watanabe superprocesses.  
However, as a result of the branching assumption (once born, individuals behave 
independently of one another), 
a population evolving according to any of these models will
either die out or grow without bound and, in so doing, it can
develop clumps of arbitrarily large density and extent. 
Our starting point here is an individual-based model of a single species 
in continuous space in which
birth, death, and establishment may all depend on local population density
as well as on spatial location,
allowing for stable populations through density-dependent feedback.

Although it is often mathematically convenient to assume that individuals follow
Brownian motion during their lifetime, 
in our model, offspring
are thrown off according to some 
spatial distribution centred on the location of the parent.   
This is particularly appropriate for modelling plant populations, in which
this dispersal of offspring around the parent is the only source of
spatial motion.

Most models do not distinguish between juveniles and adults, so,
for example, the number of adults produced by a single parent is determined
by the degree of crowding at the location of the parent. A novelty
of the models that we introduce here is that, although we shall only
follow the adult population, in formulating the dynamics of the
models we shall distinguish
between production of juveniles, which will depend upon the location of 
the adult, and their successful establishment, which will depend on the
location in which a juvenile lands. The result is that not only the absolute 
number, but also the spatial distribution
around their parent, 
of those offspring that survive to adulthood
will depend upon the local population 
density.

We shall consider three different classes of scaling limits for our model.
The first yields a class of (generalised) superprocesses in which coefficients 
governing both the spatial motion and the branching components of the process can depend
on local population density; the second is a corresponding class of deterministic
non-local differential equations; and the third are classical PDEs.
We measure local population density around a point by convolving with
a smooth kernel $\rho(\cdot)$, which may differ
for the two stages of reproduction. 
When the limit is deterministic,
it is a (weak) solution of an equation of the form
\begin{equation}
	\label{general deterministic limit}
        \partial_t \varphi_t(x)
        =
        r\left(x, \varphi_t \right)
        \DG^* \left[
            \varphi_t(\cdot)
            \gamma\big( x, \varphi_t \big)
        \right](x)
        +
        \varphi_t(x)
        F\left(x, \varphi_t \right)
        ,
\end{equation}
where $\DG^*$ is (the adjoint of) a strictly uniformly elliptic
second order differential operator, typically
the Laplacian.
We shall be more specific about the parameters below.

By replacing $\rho$ by
$\rho^\epsilon(\cdot)=\rho(\cdot/\epsilon)/\epsilon^d$, 
we can also scale the ``width'' of the 
region over which we measure local population density. 
When the population follows~(\ref{general deterministic limit}), 
we can expect that taking a second limit by scaling the kernels appearing
in $r$, $\gamma$, and $F$ in this way, so that
in the limit as $\epsilon\to 0$
interactions are pointwise, we should
recover a nonlinear PDE.
We verify that this is indeed the case in two important examples:
a special case of the porous medium equation with a logistic growth term, in which we 
retain a signal of our two stage approach to modelling in the nonlinear diffusion term
and the limiting equation takes the form 
\begin{equation}
	\label{PME1}
	\partial_t \varphi = \Delta (\varphi^2)+\varphi(1-\varphi);
\end{equation}
and a wide class of semi-linear PDEs of the form 
\begin{equation}
\label{semilinear PDE}
\partial_t \varphi = \DG^*\varphi+ \varphi F(\varphi),
\end{equation}
which includes the Fisher-KPP equation and the Allen-Cahn equation.

It is of interest to understand under what conditions we can replace the two-step limiting
process described above by one in which we simultaneously scale the 
kernels
and the other parameters in 
our population model to arrive at the PDE limit.
This is mathematically much more challenging, but we establish such 
one-step convergence in cases for which the limit is a classical 
reaction diffusion equation of the form~(\ref{semilinear PDE}) with $\DG=\Delta$
and $\rho_F$ is a Gaussian density. We allow a wide class of reaction terms, $F$, so
that the Fisher-KPP equation emerges as a special case.

We remark that such results on (one-step) convergence to reaction diffusion 
equation limits have been
achieved for a variety of interacting particle systems. The vast majority of these
results concern spin systems \comment{REFs}. However,~\cite{flandoli/huang:2021} 
obtain the Fisher-KPP equation (that is equation~(\ref{semilinear PDE}) with 
$\DG=\Delta$ and $F(\varphi)=1-\varphi$) as a scaling limit of interacting diffusions in
$\IR^d$. In that setting a major challenge (also apparent in our models), not
faced in spin systems, is 
the control of concentration of particles. We also note that the scaling 
considered in~\cite{flandoli/huang:2021} is different from that considered
below. In their language we consider only `moderate interactions', meaning that
the number of individuals in the neighbourhood over which we measure local
population density tends to infinity, whereas they also consider the situation in which
that number remains finite; furthermore, whereas we shall consider a scaling in which 
the reproduction mechanism is near-critical
and look at large timescales,~\cite{flandoli/huang:2021}
only scale the number of
particles in the system and the size of the neighbourhood over which indivduals
interact.

The history of a natural population is often only accessible indirectly, 
through patterns of genetic diversity that have been laid down; from 
genetic data, one can try to infer the genealogical trees that relate 
individuals in a sample from the population, and these have been shaped 
by its history. It is therefore of interest
to establish information about the distribution of genealogical trees 
under our population model.
To achieve this we exploit a lookdown construction. 
Lookdown constructions were first introduced in~\cite{donnelly/kurtz:1996} to provide 
a mechanism for retaining information about genealogical relationships between individuals
sampled from a population evolving according to the Moran model when passing to the 
infinite population limit.
Since then, they have been extended to a huge range of models. Of particular
relevance to our work here are the papers of~\cite{kurtz/rodrigues:2011} 
and~\cite{etheridge/kurtz:2018}, in which
lookdown constructions are provided for a wide variety of population models, including
spatially structured branching processes.

In general, even armed with a lookdown construction, calculation of relevant statistics 
of the genealogy remains a difficult question. 
However, in special
circumstances, some progress can be made. As an illustration, 
we shall consider a scenario that
has received a great deal of attention in recent years, in which 
a population is expanding into new territory as a travelling wave. 
In Section~\ref{ancestral lineages for travelling waves}
we shall consider three different (deterministic) wavefronts, moving 
across $\IR^1$, and we shall explore the position of a single ancestral 
lineage relative to that front.

Most work has focused on models that caricature the classical Fisher-KPP equation with a 
stochastic term, i.e.
$$dw=\big(\Delta w +sw(1-w)\big)dt +\sqrt{\frac{\alpha(w)}{N}}W(dt,dx),$$
where $W$ is space-time white noise, and $N$ is a measure of the 
local population density. The coefficient
$\alpha(w)$ is generally taken to be either
$w$, corresponding to a superprocess limit, or $w(1-w)$ giving a 
spatial analogue of a Wright-Fisher diffusion. 
For very large $N$, and on suitable timescales, 
starting with the pioneering work of~\cite{brunet/derrida/mueller/munier:2006},
a considerable body of evidence has been amassed to underpin the
conjecture that for this, and a wide class of 
related models, 
genealogies converge to a 
Bolthausen-Sznitman coalescent. %, e.g.~\cite{brunet/derrida/mueller/munier:2007}.
This reflects the fact that, for this equation, 
ancestral lineages become trapped in 
the wavefront, where the growth rate of the population is highest. 
Once there, they will experience 
rapid periods of coalescence 
corresponding to significant proportions of individuals in the front being 
descended from particularly reproductively successful ancestors.

If one replaces the logistic growth term of the classical
Fisher-KPP equation with a nonlinearity that reflects cooperative
behaviour in the population, such as
\begin{equation}
	\label{Birzu nonlinearity}
	wF(w)=w(1-w)(Cw-1),
\end{equation}
then, for sufficiently large $C$ (strong cooperation),
the nature of the deterministic
wave changes from ``pulled'' to ``pushed'', \cite{birzu/hallatschek/korolev:2017}.
In that setting, the genealogies will be quite different
from the Fisher-KPP case. \comment{Could refer to more recent Birzu et al paper here.}
For example, \cite{etheridge/penington:2022}
show that for a discrete space model corresponding to this 
nonlinearity with $C>2$, after suitable scaling, the genealogy of a
sample converges not to a Bolthausen-Sznitman coalescent, but to
a Kingman coalescent. 
The reason, roughly, is that ancestral lineages
settle to a stationary distribution relative to the position of the 
wavefront which puts very little weight close to the `tip' of the wave, so
that when ancestral lineages meet
it is typically at a location in which population density is
high, where no single ancestor produces a disproportionately large number of 
descendants in a short space of time.

The shape of the wave is not determined solely by the reaction term.
For example, as a result of the nonlinear diffusion, 
for suitable initial conditions, the solution to the one-dimensional 
porous medium equation with logistic growth~(\ref{PME1}) converges to a travelling
wave with a sharp cut-off; i.e., in contrast to the classical 
Fisher KPP equation, the solution at time $t$
vanishes beyond $x=x_0+ct$ for some constant wavespeed
$c>0$, \cite{kamin/rosenau:2004}.
As a first step towards understanding what we should expect in models with
nonlinear diffusion, one can ask about the position of an ancestral lineage
relative to the wavefront in the deterministic models. In 
Section~\ref{ancestral lineages for travelling waves}
we shall see that in our framework, even with logistic growth, the nonlinear diffusion 
corresponding to the porous medium equation results
in a stationary distribution that is concentrated behind the wavefront, leading 
us to conjecture that in the stochastic equation
the cooperative behaviour captured by the nonlinear diffusion will also 
result in a qualitatively different pattern of coalescence to that seen under the 
stochastic Fisher-KPP equation. Indeed, we believe that it should be feasible to 
show that in an appropriate limit one recovers a Kingman coalescent.

% % % % % % % % % % % % % %
\paragraph{Structure of the paper}

\comment{This paragraph to be rewritten last}

In this paper we study scaling limits of spatial population models,
obtaining convergence of both the population process
(i.e., the population density as a function of time,
although strictly speaking it is a measure that may not have a density)
and of lineages traced back through such a population.
We retain information about lineages as we pass to the scaling limit
by means of a lookdown construction.
First, in Section~\ref{sec: Model and main results},
we describe the model and the main results
(Theorems~\ref{thm:nonlocal_convergence}, \ref{thm:local_convergence}, and~\ref{thm:lineages}).
Next, in Section~\ref{sec:applications}, we discuss a few striking consequences of these results,
namely, X Y and Z \comment{TODO}.
In Section~\ref{sec:heuristics}, we provide heuristic explanations
of why the theorems ought to be true,
and some key ideas behind them.
In Section \ref{sec:lookdown} we define and discuss the lookdown construction.
This is followed by the technical proofs in Section~\ref{sec:proofs}.

\comment{What's in the Appendix?}

@@

Our model is one of individuals distributed across a continuous space 
which we shall take to be $\IR^d$. 
For applications, $d=1$ or $d=2$, but our main results apply more generally.
At time zero, the population is distributed over a bounded region, with 
$\bigO(N)$ individuals per unit area in that region,
so the total number of individuals will also be $\bigO(N)$.
The population changes in continuous time, and
we encode the state of the population at time $t$ by a counting measure $X(t)$,
which assigns one unit of mass to the location of each individual.

Population dynamics are controlled by three quantities,
birth ($\gamma$), establishment ($r$), and death ($\mu$),
each of which can depend on spatial location and local population density
in a way specified below.
Each individual gives birth at rate $\gamma$ to a single (juvenile) offspring,
which is dispersed according to a kernel $q(x, \cdot)$ away from the location $x$ of the parent.
We assume that $q$ is the density of a multivariate Gaussian,
allowing a nonzero mean and anisotropic variance.
Both the mean and covariance can change across space,
but do not depend on population density.
The offspring does not necessarily survive to be counted in the population:
it ``establishes'' with probability $r$,
or else it dies immediately.
Independently, each individual dies with rate $\mu$.

We aim to capture universal behaviour by passing to a scaling limit. Specifically, 
we shall take the ``density'', $N$, to infinity,
and also scale time by a factor of $\theta$,
in such a way that defining $\eta^N(t) = X(\theta t) / N$,
the process $\{\eta^N(t)\}_{t \ge 0}$
will converge to a suitable measure-valued process
as $N$ and $\theta$ tend to infinity,
with the nature of the limit depending on how they tend to infinity together.
Evidently, we shall also need to scale the dispersal kernel if we are to obtain a
nontrivial limit. We write $q_\theta(x,\cdot)$ for the density of the 
multivariate Gaussian obtained by
multiplying the mean and variance components corresponding to $q(x,\cdot)$ by $1/\theta$.

Birth, establishment, and death can depend on the location of the individual
and the local population density.
Since we would like the population density to scale with $N$,
these are functions of $X/N$, i.e.,
the counting measure with mass $1/N$ placed at the location of each individual.
First consider birth rates, defined by
a nonnegative function $\gamma(x, m) : \IR^d \times \IR_{\ge 0} \to \IR_{\ge 0}$
of location $x$ 
and local population density $m$.
Local population density is defined as the convolution of $X/N$ with 
a smooth (non-negative, integrable) kernel $\rho_\gamma(\cdot)$.
We write this convolution as $\smooth{\gamma} X/N$.
Then, when the state of the population is $X$, an individual at location $x$ 
gives birth to a single juvenile offspring at rate
$\gamma(x, \smooth{\gamma} X(x) / N)$.
Similarly, the establishment probability of an offspring at location $y$ is
is $r(y, \smooth{r} X(y) / N)$,
where $r(y, m) : \IR^d \times \IR_{\ge 0} \to [0, 1]$
and again $\smooth{r} X$ is the convolution of 
$X/N$ with the smooth kernel $\rho_r$.

We shall write $\mu_\theta(x, X/N)$ for the per-capita death rate of mature individuals
in the population.
In order for the 
population density to change over timescales of order $\theta$, we should like the net 
per capita reproductive rate 
to scale as $1/\theta$. 
In classical models, in which $r$, $\gamma$, and $\mu$ are constant, 
this quantity is simply $r\gamma -\mu$. 
Here, because production of juveniles and their establishment are
mediated by population density measured relative to different points, 
the net reproductive rate will take a more complicated form. 
In particular, the total rate of production of mature offspring by an individual at $x$ will be
\[
	\gamma\big(x,\smooth{\gamma}X(x)/N\big)\int r\big(y, \smooth{r}X(y)/N\big)q_{\theta}(x,dy).
\]
Nonetheless it will be convenient to define the death rate $\mu_\theta$ in terms of 
its deviation from $r\gamma$. 
To this end, we define
the death rate of an individual at $x$ to be
\begin{align} \label{eqn:mu_defn}
    \mu_\theta(x)
    =
    \max\left\{0, 
        r(x, \smooth{r} X(x) / N) \gamma(x, \smooth{\gamma} X(x) / N)
        - \frac{1}{\theta} F(x, \smooth{F} X(x) / N)
    \right\} ,
\end{align}
where $F(x, m) : \IR^d \times \IR_{\ge 0} \to \IR$ and
$\rho_F$ is again a smooth kernel.

Note that each of the three demographic parameters $r$, $\gamma$, $F$,
depends on local density,
measured by convolution with a smooth kernel. Each of the three kernels can be different.
As a result, death rate depends (in principle) on population densities measured in 
three different ways, 
so that we could write 
$\mu_\theta(x) = \mu_\theta(x, \smooth{\gamma} X(x) / N, \smooth{r} X(x) / N, \smooth{F} X(x) / N)$.
This may seem unnecessarily complex.
However, not only is it natural from a 
biological perspective, it also turns out to be necessary if we are
to capture nontrivial examples in our scaling results.

\begin{remark}
Although this model allows fairly general birth and death mechanisms,
there are a number of limitations.
Perhaps most obviously, to simplify the notation
individuals give birth to only one offspring at a time,
although this restriction could be easily lifted
\citep[as in Section 3.4 of][]{etheridge/kurtz:2018}.
Furthermore, individuals do not move during their lifetime,
and the age of an individual does not affect its fecundity or death rate.
Finally, there is no notion of mating
(although limitations on reproduction due to availability of mates can be incorporated into the birth rate, $\gamma$),
so the lineages we follow will be uniparental.
For these reasons, the model is most obviously applicable to bacterial populations or selfing plants,
although we do not anticipate that incorporation of these complications
will change the general picture.
\end{remark}

For each $N$ and $\theta$, the rescaled process
$\big(\eta^N_t\big)_{t\geq 0} := \big(X(\theta t)/N\big)_{t\geq 0}$, 
takes values in the space of c\`adl\`ag paths in
$\measures$ (the space of finite measures on $\IR^d$
endowed with the weak topology).
